A Simple Minibatch Acceptance Test for MCMC

### 2016-05-20 nips_2016.tex
updated lots of the theory stuff, typos etc

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\frac{N}{n}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian as $n
\to \infty$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=1\textwidth]{mh_convolution_diagram_v2}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution. We use three standard deviation values of $X_{\rm norm}$. The
    two red curves convolve to form the logistic, etc. The $y$-axis is capped at 0.5 for
    readability. This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    %\vspace{-10pt}
\end{figure}

@@

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}) we make $K$
    small, around 5-10, for speed purposes, and $n$ is usually a few hundred for the Central Limit
    Theorem to apply.  Using these $K$ values, we compute the empirical standard deviation ${\rm
    std}(\Delta')$.

@@

\textbf{Variance Preconditioning}. Returning to the discussion from Section~\ref{sec:introduction},
our test requires a variance check. Equation~\ref{eq:deconvolution} means that the variance of $X$,
which is $\pi^2/3\approx 3.29$ is an upper bound on the variances of $X_{\rm norm}$. Therefore, the
test cannot work if the variance of $X_{\rm norm}$ is too large. Since the standard deviation of $X$
is about $\sqrt{3.29}\approx 1.81$, this bounds ${\rm std}(X_{\rm norm})$ (as well as ${\rm
std}(\Delta')$). From the discussion above, we bound the standard deviation at 1.2 so that $X_{\rm
corr}$ can handle the remaining noise. Recall from Section~\ref{ssec:deltas_minibatch} that we
estimate the ${\rm std}(X_{\rm norm})$ each iteration. If the variance/standard deviation test
fails, then we can (a) skip the iteration, (b) increase the minibatch size, (c) change the proposal
to decrease step sizes, or (d) increase temperature, as we now discuss.

@@

We use some terms from standard Markov chain theory~\cite{Meyn2009}. Let the \emph{transition
kernels} of MCMC on complete and minibatch data at step $i$ be $P_i$ and $P_i'$, respectively. When
these kernels (e.g., $P_i$) are applied on a probability distribution $D_1$, they generate a new
distribution $D_2$, which we write as $P_i \circ D_1 = D_2$. Define $\pi$ to be the stationary
distribution obtained by $P_i$, and $\pi_i'$ to be the stationary distribution obtained by $P_i'$
($\pi'$ needs a subscript in this case).  Finally, we indicate the \emph{total variation distance}
between two kernels using $\| \cdot \|$, e.g., $\|P_i-P_i'\|$.

Our first result shows that the distance between $P_i$ and $P_i'$ is proportional to our estimation
error $X_{\xi}$.

\begin{lemma}\label{lem:theory1}
If $|X_\xi| < \zeta$ and $|\nabla F_X(x)| < \ell$ for all $x$, then $\|P_i-P_i'\| \le 2\zeta \ell$.
\end{lemma}

@@

\begin{theorem}\label{thm:theory3}
Assume $\pi$ satisfies $\|P_i \circ D_0 - \pi\| \leq \eta \|D_0 - \pi\|$, where $\eta \in [0, 1)$ for
all distributions $D_0$. Also assume there exists $0 < \rho_t < 1$ such that $\|P'_i(x, \cdot) -
\pi_i'\| \leq 2\rho_i$. We can get $\| P_t' \circ P_{t-1}' \circ \cdots \circ P_1' \circ D_0 - \pi
\| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s \alpha_s + \alpha_t $, where
$\alpha_i = \frac{\epsilon_i C }{1-\eta}$ and $\epsilon_i$ is the step size at iteration $i$.
\end{theorem}

@@

We run MCMC with our MH test with $m=200$. We also run this using standard full-batch MCMC
(``standard sampling'') with the standard MH test, and the method from~\cite{cutting_mh_2014}
(``adaptive sampling''). For the latter, we also use $m=200$ and increment minibatches by that
amount within an iteration if necessary. The tolerance for making a decision is $\epsilon=0.1$. To
make comparisons easier, all three use the same random walk proposer with covariance $\Sigma={\rm
diag}(0.03, 0.03)$. This is a poor proposer, but it is sufficient for our purposes as the quality of
the samples will be primarily due to the MH test. All methods are run 10000 times to collect 10000
samples.

@@

In this experiment, we apply our minibatch MH test with a Stochastic Gradient Hamiltonian
Monte Carlo (SGHMC)~\cite{sghmc_2014} proposer to sample the complete parameters for a fully
connected neural network. We use the Higgs data set from the UCI dataset~\cite{Lichman:2013}, which
is a binary classification task with one million instances, each of which are 28-dimensional. The
network we use has four layers: the first is the input, the last is the softmax, and the
intermediate ones apply the sigmoid activation unit.  We also employ dropout and batch
normalization~\cite{icml2015_ioffe15}.

For our comparison baseline, we train the same network architecture using the Adaptive Gradient
Descent method~\cite{adapGrad}. Since the frequencies of the features are quite different, we use
the same strategy as in Adaptive Gradient to rescale the gradient before applying SGHMC. We
implement our MH test and Adaptive Gradient in the open-source BIDMach
project~\cite{canny2013bidmach}.

@@

Next, we show the transition kernels at each step are bounded if we use the same proposer, i.e.
$\|P_i - P'_i\| \leq 2\zeta \ell$. We write $P_i(\theta_t \mid \theta') = P_a(\theta_t, \theta')
q(\theta'\mid \theta_t) + (1-P_a(\theta_t,\theta')) \delta_D(\theta' - \theta_t)$, where $\delta_D$
is the Dirac delta function. The total variation distance between two proposal kernels is
\begin{align*}
\|P_i - P'_i \| &= \int_{\theta'} \left| \int_{\theta_t}((P_a-P_a') q(\theta'|\theta_t) + (1-P_a - 1+P_a') \delta_D(\theta_t -\theta')) d\theta_t  \right| d\theta' \\
&= \int_{\theta'} \left| \int_{\theta_t} (P_a- P_a')(q(\theta'|\theta_t) - \delta_D(\theta_t - \theta')) d\theta_t \right|  d\theta'\\ 
&\leq 2 \zeta \ell \int_{\theta'} \left| \int_{\theta_t}(q(\theta'|\theta_t) + q(\theta_t|\theta')) d\theta_t \right| d\theta'\\
&= 2 \zeta \ell,
\end{align*}
as desired.

@@

First, by Theorem 1 in~\cite{cutting_mh_2014}, $\|P_i \circ D_0 - \pi\| \leq \eta \|D_0
- \pi\|$ and $\|P_i-P'_i\| \leq \epsilon_i C$, the norm between the stationary distribution is $\|\pi
- \pi_i'\|\leq \frac{\epsilon_i C}{1-\eta}$.

@@

For SGHMC, we use the simplified update equations~\cite{sghmc_2014}:
\begin{equation}
\Delta \theta = v, \quad \quad\Delta v = -\eta \nabla U(\theta) - \alpha v + \mathcal{N}(0, 2(\alpha
-\hat{\beta}) \eta),
\end{equation}
where $v$ represents auxiliary momentum variables, and $\nabla U(\theta)$ is the gradient of the
system. We set the hyperparameters to be $\eta_i = 0.01 /(i+1)^{0.4}$, and $\alpha=0.1$. We use the
empirical Fisher~\cite{conf/icml/AhnBW12} information $V(\theta)$ to estimate the value of
$\hat{\beta}$, so that $\hat{\beta}=\frac{1}{2}\eta V(\theta)$.

In order to control ${\rm std}(\Delta')$, we initialize the temperature at 1000, and adjust it
at iteration $i$ according to $T_i = \max\{1, 1000/(i+1)^{0.5}\}$.

>>>>

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\frac{N}{n}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian for
sufficiently large $N$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=1\textwidth]{mh_convolution_diagram_v2}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution. We use three standard deviation values of $X_{\rm norm}$. The
    two red (respectively, blue, green) curves form the logistic. The $y$-axis is capped at 0.5.
    This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    %\vspace{-10pt}
\end{figure}

@@

\begin{enumerate}[noitemsep]
    \item We first need to estimate the standard deviation of $X_{\rm norm}$. We generate $K$ values
    of $\Delta'$, each using a different random minibatch of $n$ data points, but each using the
    same $\theta_t$ and $\theta'$. In our experiments (see Section~\ref{sec:experiments}) we make $K$
    small, around 5-10, for speed purposes, and $n$ is usually 50-200.  Using these $K$ values, we
    compute the empirical standard deviation ${\rm std}(\Delta')$.

@@

\textbf{Variance Preconditioning}. Returning to the discussion from Section~\ref{sec:introduction},
our test requires a variance check. Equation~\ref{eq:deconvolution} means that the variance of $X$,
which is $\pi^2/3\approx 3.29$ is an upper bound on the variances of $X_{\rm norm}$. Therefore, the
test cannot work if the variance of $X_{\rm norm}$ is too large. The standard deviation of $X$
is about $\sqrt{3.29}\approx 1.81$, so this bounds ${\rm std}(X_{\rm norm})$ (as well as ${\rm
std}(\Delta')$). From the discussion above, we bound the standard deviation at 1.2 so that $X_{\rm
corr}$ can handle the remaining noise. Recall from Section~\ref{ssec:deltas_minibatch} that we
estimate the ${\rm std}(X_{\rm norm})$ each iteration. If the variance/standard deviation test
fails, then we can (a) skip the iteration, (b) increase the minibatch size, (c) change the proposal
to decrease step sizes, or (d) increase temperature, as we now discuss.

@@

We use some terms from standard Markov chain theory~\cite{Meyn2009}. Let the \emph{transition
kernels} of MCMC on complete and minibatch data at step $i$ be $P_i$ and $P_i'$, respectively. When
these kernels (e.g., $P_i$) are applied on a probability distribution $D_1$, they generate a new
distribution $D_2$, which we write as $P_i \circ D_1 = D_2$. Define $\pi$ to be the stationary
distribution obtained by $P_i$, and $\pi_i'$ to be the stationary distribution obtained by $P_i'$
($\pi'$ needs a subscript in this case).  Finally, we indicate the \emph{total variation distance}
between two kernels using $\| \cdot \|$.

Our first result shows that the distance $\|P_i\circ D - P_i'\circ D\| \le 4\zeta \ell$ for all
distributions $D$.

% Our first result shows that the distance between $P_i$ and $P_i'$ is proportional to our estimation
% error $X_{\xi}$.

\begin{lemma}\label{lem:theory1}
If $|X_\xi| < \zeta$ and $|\nabla F_X(x)| < \ell$ for all $x$, then $\|P_i\circ D-P_i'\circ D\| \le
4\zeta \ell \;\;\forall D$.
\end{lemma}

@@

\begin{theorem}\label{thm:theory3}
Assume $\pi$ satisfies $\|P_i \circ D_0 - \pi\| \leq \eta \|D_0 - \pi\|$, where $\eta \in [0, 1)$ for
all distributions $D_0$. Also assume there exists $0 < \rho_t < 1$ such that $\|P'_i \circ D -
\pi_i'\| \leq 2\rho_i \;\; \forall D$. We can get $\| P_t' \circ P_{t-1}' \circ \cdots \circ P_1' \circ D_0 - \pi
\| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s \alpha_s + \alpha_t $, where
$\alpha_i = \frac{\epsilon_i C }{1-\eta}$ and $\epsilon_i$ is the step size at iteration $i$.
\end{theorem}

@@

We run MCMC with our MH test using minibatch size $m=200$. We also run this using standard
full-batch MCMC (``standard sampling'') with the standard test from Equation~\ref{eq:traditional}, and the method
from~\cite{cutting_mh_2014} (``adaptive sampling''). For the latter, we also use $m=200$ and
increment minibatches by that amount within an iteration if necessary. The tolerance for making a
decision is $\epsilon=0.1$. To make comparisons easier, all three use the same random walk proposer
with covariance $\Sigma={\rm diag}(0.03, 0.03)$. This is a poor proposer, but it is sufficient for
our purposes as the quality of the samples will be primarily due to the MH test. All methods are run
10000 times to collect 10000 samples.

@@

In this experiment, we apply our minibatch MH test with a Stochastic Gradient Hamiltonian
Monte Carlo (SGHMC)~\cite{sghmc_2014} proposer to sample the complete parameters for a fully
connected neural network. We use the Higgs data set from the UCI dataset~\cite{Lichman:2013}, which
is a binary classification task with one million instances, each of which are 28-dimensional. The
network we use has four layers: the first is the input, the last is the softmax, and the
intermediate ones apply the sigmoid activation unit.  We also employ dropout and batch
normalization~\cite{icml2015_ioffe15}.  In order to control ${\rm std}(\Delta')$, we initialize the
temperature at 1000, and adjust it at iteration $i$ according to $T_i = \max\{1,
1000/(i+1)^{0.5}\}$.

For our comparison baseline, we train the same network architecture using the Adaptive Gradient
Descent method~\cite{adapGrad}. Since the frequencies of the features are quite different, we use
the same strategy as in Adaptive Gradient to rescale the gradient before applying SGHMC. We
implement our MH test and SGHMCin the open-source BIDMach project~\cite{canny2013bidmach}.

@@

Next, we show the distance of new distributions obtained by applying $P_i$ and $P_i'$ on an
arbitrary distribution $D$ once is bounded if we use the same proposer, i.e.  $\|P_i\circ D -
P'_i\circ D\| \leq 4\zeta \ell \;\; \forall D$. We write $P_i(\theta_t \mid \theta') = P_a(\theta_t, \theta')
q(\theta'\mid \theta_t) + (1-P_a(\theta_t,\theta')) \delta_D(\theta' - \theta_t)$, where $\delta_D$
is the Dirac delta function. The total variation distance between two generated distributions is
\begin{align*}
\|P_i\circ D - P'_i\circ D\| &= \int_{\theta'} \left| \int_{\theta_t}((P_a-P_a') q(\theta'|\theta_t) + (1-P_a - 1+P_a') \delta_D(\theta_t -\theta')) dD(\theta_t) \right| d\theta' \\
&= \int_{\theta'} \left| \int_{\theta_t} (P_a- P_a')(q(\theta'|\theta_t) - \delta_D(\theta_t - \theta')) dD(\theta_t) \right|  d\theta'\\ 
&\leq 2 \zeta \ell \int_{\theta'} \left| \int_{\theta_t}(q(\theta'|\theta_t) + \delta_D(\theta_t - \theta')) dD(\theta_t) \right| d\theta'\\
&\leq 2 \zeta \ell \int_{\theta'} ( D'(\theta') + D(\theta') ) d\theta'\\
&= 4 \zeta \ell,
\end{align*}
where $D'$ is the distribution generated if we always accept the proposed point. (In the last line,
we integrate two probability density functions, resulting in a value of two.) This matches our
desired bound.

@@

First, by Theorem 1 in~\cite{cutting_mh_2014}, $\|P_i \circ D_0 - \pi\| \leq \eta \|D_0
- \pi\|$ and $\|P_i\circ D-P'_i\circ D\| \leq \epsilon_i C$ for all $D$. The norm between the
stationary distributions is $\|\pi - \pi_i'\|\leq \frac{\epsilon_i C}{1-\eta}$.

@@

For SGHMC, we use the simplified update equations~\cite{sghmc_2014}:
\begin{equation}
\Delta \theta = v, \quad \quad\Delta v = -\eta \nabla U(\theta) - \alpha v + \mathcal{N}(0, 2(\alpha
-\hat{\beta}) \eta),
\end{equation}
where $v$ represents auxiliary momentum variables, and $\nabla U(\theta)$ is the gradient of the
system. We set the hyperparameters to be $\eta_i = 0.01 /(i+1)^{0.4}$, and $\alpha=0.1$. We use the
empirical Fisher~\cite{conf/icml/AhnBW12} information $V(\theta)$ to estimate the value of
$\hat{\beta}$, so that $\hat{\beta}=\frac{1}{2}\eta V(\theta)$. In order to control ${\rm
std}(\Delta')$, we initialize the temperature at 1000, and adjust it at iteration $i$ according to
$T_i = \max\{1, 1000/(i+1)^{0.5}\}$.
